%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% CMPT 435
% Lab Zero
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from: http://www.LaTeXTemplates.com
% Original author: % Frits Wenneker (http://www.howtotex.com)
% License: CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
% Modified by Alan G. Labouseur  - alan@labouseur.com, and Patrick Tyler - Patrick.Tyler1@marist.edu
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[letterpaper, 10pt]{article} 

\usepackage[english]{babel} % English language/hyphenation
\usepackage{graphicx}
\usepackage[lined,linesnumbered,commentsnumbered]{algorithm2e}
\usepackage{listings}
\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\usepackage{lastpage}
\usepackage{url}
\usepackage{xcolor}
\usepackage{titlesec}
\usepackage{xurl}

% Stolen from https://www.overleaf.com/learn/latex/Code_listing 
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}
\lstset{style=mystyle, language=c++}


\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{page \thepage\ of \pageref{LastPage}} % Page numbering for center footer
\fancyfoot[R]{}

\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
   \normalfont \normalsize 
   \textsc{CMPT 435 - Fall 2023 - Dr. Labouseur} \\[10pt] % Header stuff.
   \horrule{0.5pt} \\[0.25cm] 	% Top horizontal rule
   \huge Assignment Two -- \LaTeX ~Data\\     	    % Assignment title
   \horrule{0.5pt} \\[0.25cm] 	% Bottom horizontal rule
}

\author{Patrick Tyler \\ \normalsize Patrick.Tyler1@marist.edu}

\date{\normalsize\today} 	% Today's date.

\begin{document}

\maketitle % Print the title

%----------------------------------------------------------------------------------------
%   CONTENT SECTION
%----------------------------------------------------------------------------------------

% - -- -  - -- -  - -- -  -
\section{Linear and Binary Searches}
\subsection{Linear Search}
A linear search is one of the most flexible search algorithms. The only requirement of this search is that
the collection of elements can be traversed and checked for equality. This algorithm simply goes through
each element and checks for equality. This algorithm works in linear time, $O(n)$, because, in the worse
case, it checks every element for equality. \\
\newline
\noindent
The following code implementation loops through an array keeping track of each iteration and returning
once the element is found or all items are traversed.
\lstinputlisting[linerange={39-52}, firstnumber=39]{main.cpp}

\subsection{Binary Search}
For a binary search to be applicable, the collection must be sorted (therefore of comparable elements)
and random access. An array of sorted elements is the most common data type in which a binary search
can be done on. An array is random access because an element can be accessed at any time with its
given index.\\
\vspace{.25cm}
\hrule
\vspace{.25cm}
\noindent
% I could not think of anything better my fault
There's no joke about binary searches that is funny. Not even \textit{a bit.}\\
\hrule
\vspace{1cm}
\noindent
This algorithm works with the following process: (1) picking the midpoint of its bounds;
(2) check equality between midpoint value and target if equal, element is found at search is complete;
(3) check whether midpoint value is greater/ lesser than target then set lower / upper bound to
midpoint -1 / +1 and repeat to step 1 if the upper bound does is not equal or lower than the lower bound. 
How the bounds change also depends on if the array is in ascending or descending order.
The following is a code example of binary search on an ascending array.
\lstinputlisting[linerange={55-71}, firstnumber=55]{main.cpp}
\section{Hash Table Search}
\subsection{Overview}
For a hash table search to be implemented the data has to be stored in well... a hash table. A hash table maps
a value to a key with a hash function. In an array implementation the key would be an index. A hash function
should be able to hash an indefinite number of inputs depending only constrained by the data type itself. 
Ideally this mapping is 1 to 1, but practically mapping an indefinite number of inputs to a key also requires
an indefinite amount of space to store the value at each given key. Thus, there may be collisions
where values map to the same key; this will have to be dealt with through storing a collection at each key
which can dynamically increase in size such as a linked list. The search process of a hash table is
hashing the target value then going to its collection and finding or not finding (usually through a linear search)
the target in the chain.
\subsection{Absurdity of a Perfect Hash Function}
A perfect hash function maps every valid value as a unique key. As an exercise, let's see how many keys
are needed to create a perfect hash function for 8 bit ascii strings. Each ascii character
has $2^8$ or 256 different possibilities. So, if the hash function only hashes
a single ascii character it will need 256 different keys. If the valid input is extended each
additional character will multiply the number of different keys needed by 256 because there will be an
additional 256 valid strings for each already existing combination. To be able to hash just 16 character
strings there will need to be $256^{16}$ or $2^{128}$ keys. Each key will need 1 bit
to store whether it exists in
the hash table (it will not need to store the string itself because it is unique).
In total, this hash table will be $2^{128}$ bits. This is about $4*10^{16}$ \textit{zettabytes}. The amount
of storage -- \textit{not even working memory} -- in all of the world's computers is a rounding error compared to
this number. So it is safe to say: prefect hashes are impractical for hash tables.
\vspace{.25cm}
\hrule
\vspace{.25cm}
% I am really reaching to find jokes for these topics
\noindent
Guy: \textit{Hey girl, are you are perfect hashing function?}\\
Girl: \textit{What?}\\
Guy: \textit{Because you're impossible to find and entirely unique.}\\
Girl: \textit{...}\\
\hrule
\vspace{1cm}
\newpage
\subsection{Sample Hash Function}
The following is an example of an imperfect hash that has hashSize different keys (250 for later examples). This maps any string
into an integer from 0 to hashSize-1. It does this through summing the ascii values (treating all alphabetical characters as uppercase)
and multiplying by a large prime and the finding the reminder after dividing by the hashSize. This modulo operation ensures any arbitrarily
large product is converted to an integer in the correct bounds. Multiplication by a large prime is done to spread the results across the keys.
This is a common technique as many hashing methods have trends such that if the input values also have trends
it can result in clustered keys.
\lstinputlisting[linerange={143-157}, firstnumber=143]{main.cpp}

\subsection{Sample Add Implementation}
A hash table can be implemented with an array which is the same length as the amount of keys. In this implementation
both an array for the heads of the linked list and the tails is created to ensure constant time $O(1)$ adding as well as 
maintaining relative insertion order even if
the chain of the a specific key grows indefinitely large. Other implementations of adding to a hash table may have a 
factor of the chain length of each index; a possible advantage could be ensuring each value is only stored once.
This factor is often called the load factor or $\alpha$; it can
be calculated by dividing the amount of elements added to the hash table by the amount of keys.
In the this case adding would be $O(1 + \alpha)$. \\
\newline
\noindent
\newpage
The following code adds to the linked list at the tail much like an enqueue from a queue.
\lstinputlisting[linerange={123-129}, firstnumber=123]{main.cpp}
\lstinputlisting[linerange={159-195}, firstnumber=159]{main.cpp}
\section{Analysis of Searches} 
% Idk how to make this go on the next page. The table is big that the section 
%   title and table title do not fit on the same page by like one line
\newpage
\begin{center}
    \bf{Linear Search $O(n)$}\\
    \begin{tabular}{|| c c ||}
    \hline
    \bf{Comparisons} & \bf{Name} \\
    \hline
    575 & Sword of the Kauhns \\
    65  & Blood Spear +2\\
    234 & Fire Stones\\
    370 & Manual of quickness of action +3\\
    465 & Ring of Telepathic Bonds\\
    224 & Feather token, fan\\
    423 & Pipes of haunting\\
    380 & Metal-morphic Hammer\\
    233 & Fire Javelin\\
    162 & Cup of Change\\
    96  & Bracers of armor +6\\
    66  & Bloodstone Ring\\
    294 & Helm\\
    316 & Hunter's Bow\\
    253 & Gems of Darkness\\
    570 & Sword of Life\\
    187 & Dream Weaver\\
    31  & Armatha's long sword\\
    395 & Necklace of fireballs type III\\
    435 & Potion of Liquefaction\\
    620 & Troll Jar\\
    581 & Teleport Ribbon\\
    442 & Quicksilver Amulet\\
    421 & Phylactery of faithfulness\\
    177 & Dior Droid\\
    101 & Breast plate of the champion\\
    225 & Feather token, swan boat\\
    364 & Manual of gainful exercise +2\\
    326 & Ioun stone, orange\\
    602 & Tome of leadership and influence +3\\
    569 & Sword of Kings\\
    40  & Backhand\\
    14  & Amulet of mighty fists +2\\
    94  & Bracers of armor +2\\
    237 & Flail of Armor Disruption\\
    606 & Tome of understanding +2\\
    116 & Casters Aid\\
    413 & Pearl of power, 9th-level spell\\
    300 & Helm of underwater action\\
    213 & Exploding Caltrops\\
    58  & Belt of Keeping\\
    654 & Wings of flying\\
    \hline
    Average $\approx$ 308.26 & \\
    \hline
    \hline
    \end{tabular}
\end{center}
% average is 308.26
\newpage
\begin{center}
    \bf{Binary Search $O(log_2n)$}\\
    \begin{tabular}{|| c c ||}
    \hline
    \bf{Comparisons} & \bf{Name} \\
    \hline
    9  & Sword of the Kauhns\\
    9  & Blood Spear +2\\
    9  & Fire Stones\\
    10 & Manual of quickness of action +3\\
    8  & Ring of Telepathic Bonds\\
    10 & Feather token, fan\\
    8  & Pipes of haunting\\
    9  & Metal-morphic Hammer\\
    7  & Fire Javelin\\
    10 & Cup of Change\\
    9  & Bracers of armor +6\\
    10 & Bloodstone Ring\\
    9  & Helm\\
    10 & Hunter's Bow\\
    10 & Gems of Darkness\\
    9  & Sword of Life\\
    9  & Dream Weaver\\
    9  & Armatha's long sword\\
    5  & Necklace of fireballs type III\\
    9  & Potion of Liquefaction\\
    9  & Troll Jar\\
    9  & Teleport Ribbon\\
    7  & Quicksilver Amulet\\
    7  & Phylactery of faithfulness\\
    9  & Dior Droid\\
    9  & Breast plate of the champion\\
    8  & Feather token, swan boat\\
    9  & Manual of gainful exercise +2\\
    10 & Ioun stone, orange\\
    9  & Tome of leadership and influence +3\\
    8  & Sword of Kings\\
    10 & Backhand\\
    10 & Amulet of mighty fists +2\\
    9  & Bracers of armor +2\\
    10 & Flail of Armor Disruption\\
    9  & Wings of flying\\
    8  & Tome of understanding +2\\
    9  & Casters Aid\\
    8  & Pearl of power, 9th-level spell\\
    10 & Helm of underwater action\\
    9  & Exploding Caltrops\\
    10 & Belt of Keeping \\
    \hline
    Average = 8.90 & \\
    \hline
    \hline
    \end{tabular}
    % average is 8.90
\end{center}
\newpage
\begin{center}
    \bf{Hash Table Search $O(1 + \alpha)$}\\
    \begin{tabular}{|| c c ||}
    \hline
    \bf{Comparisons} & \bf{Name} \\
    \hline
    3  & Sword of the Kauhns\\
    1  & Blood Spear +2\\
    3  & Fire Stones\\
    3  & Manual of quickness of action +3\\
    2  & Ring of Telepathic Bonds\\
    1  & Feather token, fan\\
    6  & Pipes of haunting\\
    1  & Metal-morphic Hammer\\
    3  & Fire Javelin\\
    3  & Cup of Change\\
    1  & Bracers of armor +6\\
    2  & Bloodstone Ring\\
    2  & Helm\\
    3  & Hunter's Bow\\
    1  & Gems of Darkness\\
    2  & Sword of Life\\
    3  & Dream Weaver\\
    1  & Armatha's long sword\\
    3  & Necklace of fireballs type III\\
    3  & Potion of Liquefaction\\
    4  & Troll Jar\\
    5  & Teleport Ribbon\\
    2  & Quicksilver Amulet\\
    2  & Phylactery of faithfulness\\
    2  & Dior Droid\\
    1  & Breast plate of the champion\\
    3  & Feather token, swan boat\\
    5  & Manual of gainful exercise +2\\
    1  & Ioun stone, orange\\
    2  & Tome of leadership and influence +3\\
    2  & Sword of Kings\\
    1  & Backhand\\
    1  & Amulet of mighty fists +2\\
    1  & Bracers of armor +2\\
    1  & Flail of Armor Disruption\\
    2  & Tome of understanding +2\\
    1  & Casters Aid\\
    3  & Pearl of power, 9th-level spell\\
    2  & Helm of underwater action\\
    1  & Exploding Caltrops\\
    1  & Belt of Keeping\\
    2  & Wings of flying\\
    \hline
    Average $\approx$ 2.19 & \\
    \hline
    \hline
    \end{tabular}
\end{center}
\newpage
\noindent
The comparison average for each of the searches agrees with its respective time complexity
for input size n of 666. For linear search's complexity, $O(n)$, implies the comparisons will grow linearly with
the input size. This is true as the average is 308.26 or about half of the 666 length input. Binary
search's complexity, $O(log_2n)$, implies the comparisons will grow logarithmically with the input size.
This is true as the average is 8.90 which is a little less than $\lceil log_2666\rceil$.
The ceiling is taken because the comparison start counting after the first equality check of midpoint;
if the comparisons were counted only for greater or greater than or less than comparisons, the rare 
case in which the target is found on the first equality check, it would count 0 comparisons which
seems nonsensical. Regardless of how comparisons, the trend will still be logarithmic. Finally,
the hash table's complexity, $O(1 + \alpha)$, implies the comparisons will grow with the load factor.
The load factor for an input size of 666 and a key amount of 250 is $666/250\approx2.66$. The load factor
and average comparisons are about the same which is expected.\\
\newline
Each search method has advantages and drawbacks: linear search has low overhead and a lot of
flexibility with the data structure the collection is stored in; binary search only works on sorted,
random access collections, but the overhead is low and the comparisons scale logarithmically;
hash tables have relatively more overhead (especially if operations such as iteration are needed),
but they can significantly reduce the amount of comparisons by the load factor.
\vspace{.25cm}
\hrule
\vspace{.25cm}
% I am really reaching to find jokes for these topics
\noindent
Girl: \textit{I think we need to take some space from each other.}\\
Guy: \textit{Great! Then we'll spend more time together.}\\
Girl: \textit{What?}\\
Guy: \textit{It's a classic tradeoff: space versus time. "Taking space" will allow us to spend more time together}.\\
\textit{* 10 second pause *}\\
Girl: \textit{I've been cheating on you for 5 years.}\\
\hrule
\vspace{1cm}
\newpage
\section{Display Listings}
Javascript developers frontend developers seem to want a framework or library for everything.
Here is an example of a very small package for padding many javascript developers used as a 
dependency and it ended up breaking their website:
\url{https://qz.com/646467/how-one-programmer-broke-the-internet-by-deleting-a-tiny-piece-of-code.}
\lstinputlisting[linerange={80-88}, firstnumber=80]{main.cpp}
\vspace{.25cm}
\hrule
\vspace{.25cm}
\noindent
What is the difference between a javascript developer and an Amazon warehouse employee? \\ 
\textit{Nothing they're both just package managers.}\\
\hrule
\vspace{1cm}
\noindent
The rest of the display code is self-documenting and can be found in the online resources then assignment 2 then main.cpp.

\end{document}